---
title: "Exploratory Data Analysis"
author: "Sushmita V Gopalan"
fig_width: 1
fig_height: 1 
output:
  xaringan::moon_reader:
    css: ["default", "rladies", "rladies-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

# Hi, I'm Sushmita!

--
* decision science at Grubhub


--
* co-organizer at R-Ladies Chicago 


--
* formerly
   - data scientist at Northwestern Neighbourhood and Network Initiative
   - computational social science at University of Chicago 
   - experiments at Institute for Financial Management and Research 
   - economics at IIT Madras

---

# Today, we're going to 


--
* learn what exploratory data analysis is 


--
* talk about what EDA should leave us poised to do 


--
* understand some general approaches that can be applied to any datasets 


--
* see an example of EDA on a dataset of housing prices in Bangalore


--
* if time and infrastructure permit, work through an example with the famed `iris` dataset 


---

# When you first encounter a new dataset

--
* just *look* at it


--
* how many observations? how many columns?


--
* which variables are numerical, categorical, logical?


--
* do any variables need reclassifying? 


--
* are there missing observations?


--
* does it pass a initial sniff test?

---

# What is exploratory data analysis?

--
* first step to any analysis


--
* first set of things you do to get acquainted with a new dataset 


--
* iterative process of generating questions about your data 


--
* generate questions 


--
* get your hands dirty 


--
* look for patterns in broad strokes 


---

# Housekeeping

```{r message=FALSE, warning=FALSE}

library(tidyverse) # general data wrangling
library(ggplot2) # visualizations
library(tidylog) # track changes to your data
library(corrr) # calculate correlations
library(flipbookr)

theme_set(theme_minimal())
```

---

# Today's [dataset](https://www.kaggle.com/architsingh15/bengaluru-house-prices) - Housing Prices in Bangalore 

```{r message = FALSE}
df <- read_csv('housing_prices.csv')
glimpse(df)
```


---

#  View your data

```{r message=FALSE, warning=FALSE}
df %>% 
  head()
```

---

# Basic Cleaning 

```{r message=FALSE, warning=FALSE}
df <- df %>% 
  mutate(bedrooms = parse_number(size)) %>% 
  mutate(total_sqft = parse_number(total_sqft))  %>% 
  mutate(bedroom_cat = case_when(bedrooms < 6 ~ as.character(bedrooms),
                                 TRUE ~ '6+'),
         bedroom_cat = as.factor(bedroom_cat))

glimpse(df)
```

---

# Summary

```{r message=FALSE, warning=FALSE}
summary(df)
```

---

# Missing Values 

```{r geoms, message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
map_df(df, ~ sum(is.na(.))) %>%  
    t() %>%
    as.data.frame() %>%
    rename(n_NA = V1) %>% 
    rownames_to_column('variable') %>% 
    mutate(pct_NA = round(n_NA*100/nrow(df),2)) %>% 
    arrange(-pct_NA) 
```
---
# Missing Values 

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  count(balcony)
```

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  mutate(missing = ifelse(is.na(society), TRUE, FALSE)) %>% 
  group_by(missing) %>% 
  summarise(avg = mean(price))
```

---

# One Variable

* continuous? discrete?

--

* how is it distributed?

--

* are there outliers?

--

---


# Continuous Variables

* summary statistics 

* boxplots

* histograms

* smoothed histogram

```{r}
df %>% 
  select_if(is.numeric) %>% 
  head()
```

---
# Five Number Summaries

* Min
* First quartile
* Median
* Third quartile
* Maximum

```{r}
five_number_summary <- function(df, var){
df %>% 
  drop_na({{var}}) %>% 
  select({{var}}, bath) %>% 
  summarise(min = min({{var}}),
         q1 = quantile({{var}}, 0.25),
         median = quantile({{var}}, 0.5),
         q3 = quantile({{var}}, 0.75),
         max = max({{var}})) 
}

```

---

# Five Number Summaries

```{r, message=FALSE}
five_number_summary(df, total_sqft)
five_number_summary(df, bath)
five_number_summary(df, balcony)
```

---


## Five Number Summaries

```{r message=FALSE}
five_number_summary(df, price)
five_number_summary(df, bedrooms)
```

---

## Boxplots

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  ggplot() + 
  geom_boxplot(aes(x = price))
```

---

## Boxplots - remove extreme outliers? 

`x` is considered an 'extreme outlier' if it is more than 3*IQR away from the mean 

```{r}
remove_outliers <- function(df, var){
  
  IQR <- IQR(df %>% pull({{var}}))
  mean_value <- mean(df %>% pull({{var}}))
  
  df <- df %>% 
    filter({{var}} > (mean_value - 3*IQR)) %>% 
    filter({{var}} < (mean_value + 3*IQR))
  return(df)
}
```

---

## Boxplots - remove extreme outliers? 

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  remove_outliers(price) %>% 
  ggplot() + 
  geom_boxplot(aes(x = price)) +
  ggtitle("Price without extreme outliers")
```
---


## Boxplots - remove moderate outliers? 

`x` is considered an 'extreme outlier' if it is more than 3*IQR away from the mean 

```{r}
remove_outliers <- function(df, var, threshold){

  IQR <- IQR(df %>% pull({{var}}))
  mean_value <- mean(df %>% pull({{var}}))

  df <- df %>% 
    filter({{var}} > (mean_value - threshold*IQR)) %>% #<<
    filter({{var}} < (mean_value + threshold*IQR)) #<<
  return(df)
}
```
---

## Boxplots - remove moderate outliers? 
```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  remove_outliers(price,1.5) %>% 
  ggplot() + 
  geom_boxplot(aes(x = price)) +
  ggtitle("Price without moderate outliers")
```

---

## Histograms

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  ggplot() + 
  geom_histogram(aes(x = price))
```
---

## Histograms
```{r fig.width=5, fig.height=5}
df %>% 
  ggplot() + 
  geom_histogram(aes(x = price), binwidth = 2) #<<
```

---

## Density 

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  ggplot() + 
  geom_density(aes(x = price))
```


---
# Continuous Variables 

```{r message=FALSE, warning=FALSE, fig.width=7, fig.height=7, echo=FALSE}
c1 <- df %>% 
  remove_outliers(total_sqft, threshold = 3) %>% 
  ggplot() +
  geom_histogram(aes(x = total_sqft), binwidth = 50) +
  ggtitle("Square Footage")
c2 <- df %>% 
  drop_na(balcony) %>% 
  #remove_extremes(balcony) %>% 
  ggplot() +
  geom_histogram(aes(x = balcony)) +
  ggtitle("Balconies")
c3 <- df %>% 
  drop_na(bath) %>% 
  #remove_extremes(bath) %>% 
  ggplot() +
  geom_histogram(aes(x = bath)) +
  ggtitle("Bathrooms")
c4 <- df %>% 
  drop_na(bedrooms) %>% 
  #remove_extremes(bedrooms) %>% 
  ggplot() +
  geom_histogram(aes(x = bedrooms)) +
  ggtitle("Bedrooms")
gridExtra::grid.arrange(c1, c2, c3, c4, nrow = 2)
```

---
# Discrete Variables

## Unique Values

```{r message=FALSE, warning=FALSE}
df %>% 
  summarise(n_area = n_distinct(area_type),
            n_availability = n_distinct(availability),
            n_location = n_distinct(location),
            n_society = n_distinct(society)) %>% 
  t() %>% as.data.frame() %>% 
  rename('n_unique' = 'V1')
```

---
# Tables 
```{r message=FALSE, warning=FALSE}
df %>% 
  count(area_type)
```
--
```{r message=FALSE, warning=FALSE}
df %>% count(balcony)
```

---
# Bar Charts

```{r message=FALSE, warning=FALSE}
plot_bar <- function(df, var){
  df %>% 
  count({{var}}) %>% 
  mutate(pct = n/nrow(df)) %>% 
  ggplot(aes(x = {{var}}, y = pct)) + 
  geom_bar(stat = 'identity') +
  geom_text(aes(label = scales::percent(pct)), vjust = -0.5)
}

```
--
```{r message=FALSE, warning=FALSE, fig.width=4, fig.height=4}
plot_bar(df, area_type)
```

---

```{r message=FALSE, warning=FALSE, fig.width=4, fig.height=4}
df %>% 
  plot_bar(balcony)
```
---

# A mess

```{r message=FALSE, warning=FALSE, fig.width=7, fig.height=7}
df %>% 
  plot_bar(society)
```

---

## Two variables  - both are continuous

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
p1 <- df %>% 
  ggplot(aes(x = total_sqft, y = price)) + 
  geom_point() +
  geom_smooth() +
  ggtitle("Price vs Square footage")
p1
```

---

## Two variables
```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
p2 <- df %>% 
  remove_outliers(total_sqft, 3) %>% 
  ggplot(aes(x = total_sqft, y = price)) + 
  geom_point() +
  geom_smooth() +
  ggtitle("Price vs Square footage without sqft outliers")
p2
```

---

## Two variables
```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=4}
p3 <- df %>% 
  remove_outliers(price, 3) %>% 
  ggplot(aes(x = total_sqft, y = price)) + 
  geom_point() +
  geom_smooth() +
  ggtitle("Price vs Square footage without price outliers")
p3
```

---

## Two variables

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=4}
p4 <- df %>% 
  remove_outliers(price, 3) %>% 
  remove_outliers(total_sqft, 3) %>%
  ggplot(aes(x = total_sqft, y = price)) + 
  geom_point() +
  geom_smooth() +
  ggtitle("Price vs Square footage without either outliers")
p4
```

---
```{r message=FALSE, warning=FALSE, fig.width=7, fig.height=7}
gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2)
```

---
## Outcome vs continuous variables

```{r message=FALSE, warning=FALSE, fig.width=7, fig.height=7, echo=FALSE}

g1  <- df %>% 
  drop_na(price, bedrooms) %>% 
  remove_outliers(price,3) %>% 
  ggplot(aes(x = bedrooms, y = price)) + 
  geom_point() +
  geom_smooth()

g2  <- df %>% 
  drop_na(price, balcony) %>% 
  remove_outliers(price,3) %>% 
  ggplot(aes(x = balcony, y = price)) + 
  geom_point() +
  geom_smooth()

g3  <- df %>% 
  drop_na(price, bath) %>% 
  remove_outliers(price,3) %>% 
  ggplot(aes(x = bath, y = price)) + 
  geom_point() +
  geom_smooth()

g4  <- df %>% 
  drop_na(price, total_sqft) %>% 
  remove_outliers(price,3) %>% 
  ggplot(aes(x = total_sqft, y = price)) + 
  geom_point() +
  geom_smooth()

gridExtra::grid.arrange(g1,g2,g3,g4)

```

---

## Correlations

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(corrr)

plot_lower_diagonal <- function(df){
numeric_df <-  df %>% 
  select_if(is.numeric) %>% 
  replace(., is.na(.), 0)
  
correlations <- round(cor(numeric_df),1)
ggcorrplot::ggcorrplot(correlations,
           hc.order = TRUE,
           type = "lower",
           outline.color = "white") + 
  theme(axis.text.x = element_text(angle = 90)) +
  xlab(NULL) +
  ylab(NULL) +
  labs(fill = 'Correlation') +
  ggtitle("Correlations across autodispatch metrics")
}

get_correlations <- function(df){
  df %>% 
  select_if(is.numeric) %>% 
  replace(., is.na(.), 0) %>% 
  correlate() %>% 
  stretch() %>% 
  rename(var1 = x,
         var2 = y)}

plot_lower_diagonal(df)
```

---
## Continuous vs Discrete - grouping

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}

df %>% 
  remove_outliers(price,3) %>% 
  ggplot() + 
  geom_density(aes(x = price, group = area_type, colour = area_type))
```

---
## Continuous vs Discrete - faceting

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  remove_outliers(price,3) %>% 
  ggplot() + 
  geom_density(aes(x = price)) +
  facet_wrap(vars(area_type))
```

---
## Continuous vs Discrete - grouped boxplots

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  remove_outliers(price,3) %>% 
  ggplot() + 
  geom_boxplot(aes(x = area_type, y = price))
```

---

# Continuous vs Discrete - grouped summary statistics

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  group_by(location) %>% 
  summarize(average_price = mean(price),
            median_price = median(price)) %>% 
  arrange(-average_price) %>% 
  top_n(5)
```

---
# Continuous vs Discrete - grouped summary statistics

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  group_by(area_type) %>% 
  summarize(average_price = mean(price),
            median_price = median(price)) %>% 
  arrange(-average_price) %>% 
  top_n(5)
```

---
# Continuous vs cut variable

```{r message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
df %>% 
  ggplot() + 
  geom_boxplot(aes(x = bedroom_cat, y = price))
```

---
# What have we learned? 

--
* There are probably too many societies to meaningfully model 


--
* Houses with `plot_area` are generally more expensive that the other categories of `area_type`


--
* Prices are more variable in homes with more bedrooms


--
* Larger houses are more expensive overall


--
* Number of bedrooms is highly correlated with the number of bathrooms


---
# What questions do we have?

--
* Can location be reclassified?


--
* Are houses that are not in societies different from those that are in? Missing not at random?


--
* How are location, size, and price related to each other?

--
* What's the deal with `availability`?

--

---
# Remember

--
* YOU are your audience


--
* You are exploring, not explaining 


--
* If it needs to be done more than once, write a function 


--
* Resist the urge to hard code 


--
* Don't spend time making your graphs pretty


--
* Start developing a body of functions that you typically use 


---
# Thank you so much!

You can find me at 


@SushGopalan on Twitter


sushmitavgopalan16 on Github


www.sushmitagopalan.com
