---
outputs:
- Reveal
title: Hugo & hugodown
hidden: true
layout: list
weight: 1
output: hugodown::md_document
countdown: true
---

# Hugo (blog/hugo)down

A powerful static generator, usable with R Markdown.

---

# [Hugo](https://gohugo.io/)

Powerful and fast static generator

Only an .exe to install :tada:

---

# [blogdown](https://github.com/rstudio/blogdown/)

[R :package:](https://github.com/rstudio/blogdown/)

* R Markdown output formats

* Handy helpers, possibility to create custom workflows to process your website as you see fit

* RStudio add-in to create posts

Active development, many users and educators.

---

# [hugodown](https://hugodown.r-lib.org/)

[R :package:](https://hugodown.r-lib.org/)

* An R Markdown output format

* Handy helpers

Experimental. Markdown first. I use it for this website.

---

# From Rmd to website

Under the hood

```{=html}
{{<mermaid align="left">}}
graph LR;
    A[Rmd] --> |"R ( blogdown/hugodown :package:) </br> & Pandoc"| B{md}
    B --> |"Hugo (Goldmark, Chroma)"| C[HTML]
{{< /mermaid >}}
```

<small>Inspired by [Emi Tanaka's post](https://emitanaka.org/r/posts/2018-12-12-scientific-and-technical-blogging-radix-vs-blogdown/)</small>

---

# From Rmd to website

What you do

```{=html}
{{<mermaid align="left">}}
graph LR;
    A[Rmd] --> |":large_blue_circle: knit button"| B{md}
    B --> |"hugo build (locally or cloud)"| C[HTML]
{{< /mermaid >}}
```

---
---

:mountain_cableway: Time for a demo!

[Notes on the course website](/hugo/demo/)

---

Scientific Rmd Blog Checklist

* [x] R Markdown
* [x] Syntax highlighting
* [x] Modern
* [x] .bib
* [?] Citation for posts (possible but custom layout)
* [x] Equations

---

# Sustainability: blogdown

Created by:

```{r, echo=FALSE, results="asis"}
glue::glue_collapse(
  trimws(
  gsub("<.*>", "",
  desc::desc_get_author(file = find.package("blogdown"), role = "aut")
  )
  ), 
  sep = ", ")
```

Used for [Alison Hill's website](https://alison.rbind.io), in particular.

[Active development](https://github.com/rstudio/blogdown/).

---

# Sustainability: hugodown

Created by:

```{r, echo=FALSE, results="asis"}
glue::glue_collapse(
  trimws(
  gsub("<.*>", "",
  desc::desc_get_author(file = find.package("hugodown"), role = "aut")
  )
  ), 
  sep = ", ")
```

Used for [tidyverse.org](https://tidyverse.org), in particular.

[Off and on development](https://github.com/r-lib/hugodown/).

---

# Limitations?

* blogdown docs are outdated (book update planned). Markdown output not a first-class citizen yet.

* hugodown is a minimal WIP package.

* Hugo changes a lot (but both blogdown and hugodown protect your projects from that)

---

# So blogdown or hugodown?

blogdown is more mainstream, ultimately it will have more features and docs.

hugodown is more for DIY minimal workflows.

---

# Further resources

[Listed on the course website](/hugo/further-resources/) :ledger:

---

# Questions, comments?

Write them in the chat!

---

# Time for a break  :tea:

<!--html_preserve-->

<div id="timer_hugo" class="countdown" style="top:100;left:0;" data-warnwhen="0">

<code class="countdown-time"><span class="countdown-digits minutes">05</span><span class="countdown-digits colon">:</span><span class="countdown-digits seconds">00</span></code>

</div>

<!--/html_preserve-->