---
title: "Exercises BLR-R Ladies | WAQSA Workshop"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Source
These exercises are borrowed and adapted from the [R for Data Science book](https://r4ds.had.co.nz/) and Thomas Pedersen's workshop on GGPLOT2. The latter is available on [youtube](https://www.youtube.com/watch?v=0m4yywqNPVY) and a great resource if you want to explore further.   

### This document
This document contains all the code for exercises. The
document is an RMarkdown document which means that it can be compiled, along 
with the code chunks thus executing and capturing the output of the code within
the document. To read more about RMarkdown see the website for the package, as
well as the [Get Started](https://rmarkdown.rstudio.com/lesson-1.html) guide. 

### Dependencies
List of required packages is below. If library(package name) throws an error, install the package first using `install.packages("package name")`. Go ahead and run the chunk below by pressing the green button on the right. This will load all the required libraries. 

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

### Your first ggplot
We will explore the `mpg` dataset and make few basic plots.This dataset contains a subset of the fuel economy data that the EPA makes available on http://fueleconomy.gov. It contains only models which had a new release every year between 1999 and 2008 - this was used as a proxy for the popularity of the car. This data frame has 234 rows and 11 variables.


```{r}
data("mpg")
# Basic scatterplot
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy)) + 
  geom_point()
# Mapping can be given both as global (in ggplot()) or per layer
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

#Different ways of writing the same code
# mapping is understood and we don't need to state it explicitly

```

If an aesthetic is linked to data it is put into `aes()`

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class))

## Can we color subset of the data?
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class=="2seater"))
```

If you simple want to set color to a single value, placed it outside of `aes()`

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy),
             colour = 'purple')
```

#### Exercise

Modify the code below to make the points larger triangles and slightly transparent.
See `?geom_point` for more information on the point layer.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy))
```

Hint : transparency is controlled with `alpha`, and shape with `shape`
Hint 2: think about if the geometric features need to be inside or outside aes()

* * *

Using the `mpg` dataset draw a line chart, a boxplot, and a histogram


### Stat

Every geom has a stat. This is why new data (`count`) can appear when using 
`geom_bar()`.

```{r}
data("mpg")
ggplot(mpg) + 
  geom_bar(aes(x = class))
```

The stat can be overwritten. If we have precomputed count we don't want any 
additional computations to perform and we use the `identity` stat to leave the data alone

```{r}
library(dplyr)
mpg_counted <- mpg %>% 
  count(class, name = 'count')
ggplot(mpg_counted) + 
  geom_bar(aes(x = class, y = count), stat = 'identity')
```

Most obvious geom+stat combinations have a dedicated geom constructor. The one above is available directly as `geom_col()`

```{r}
ggplot(mpg_counted) + 
  geom_col(aes(x = class, y = count))
```

Values calculated by the stat is available with the `after_stat()` function 
inside `aes()`. You can do all sorts of computations inside that.

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class, y = after_stat(100 * count / sum(count))))
```

###Exercises
What does geom_col() do? How is it different from geom_bar()?

* * *

We learnt that `geom_*()` and `stat_*()` are interchangeable. Can you look at `?geom_bar()` and figure out which stat it uses as default. Modify the code below to use that stat directly instead

```{r}
ggplot(mpg) + 
  geom_bar(aes(x = class))
```

* * *

Use `stat_summary()` to add a red dot at the mean `hwy` for each group

```{r}
ggplot(mpg) + 
  geom_jitter(aes(x = class, y = hwy), width = 0.2)
```

Hint: You will need to change the default geom of `stat_summary()`

***

In our proportion bar chart, we need to set group = 1. Why? In other words what is the problem with these two graphs?
```{r}
p1<- ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop)))

p2<-ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))

p1
p2
```


### Scales
Scales define how the mapping you specify inside `aes()` should happen. All 
mappings have an associated scale even if not specified.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class))
```

Customize the color scale by specifying it separately. All scales follow the same naming convention. 
```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual')
```

Positional mappings (x and y) also have associated scales.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  scale_x_continuous(breaks = c(3, 5, 6)) + 
  scale_y_continuous(trans = 'log10')
```

#### Exercises
Use `RColorBrewer::display.brewer.all()` to see all the different palettes from
Color Brewer and pick your favourite. Modify the code below to use it. 

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual')
```

* * *

Modify the code below to create a bubble chart (scatterplot with size mapped to
a continuous variable) showing `cyl` with size. Make sure that only the present 
amount of cylinders (4, 5, 6, and 8) are present in the legend.

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class, size=cyl)) + 
  scale_colour_brewer(type = 'qual')
```

Hint: The `breaks` argument in the scale is used to control which values are
present in the legend.

Explore the different types of size scales available in ggplot2. Is the default
the most appropriate here?

* * *

Modify the code below so that colour is no longer mapped to the discrete `class`
variable, but to the continuous `cty` variable. What happens to the guide (legend)?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class, size = cty))
```

* * *

The type of guide can be controlled with the `guide` argument in the scale, or 
with the `guides()` function. Continuous colours have a gradient colour bar by 
default, but setting it to `legend` will turn it back to the standard look. What 
happens when multiple aesthetics are mapped to the same variable and uses the 
guide type?

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = cty, size = cty))
```

### Facets
The facet defines how data is split among panels. The default facet 
(`facet_null()`) puts all the data in a single panel, while `facet_wrap()` and
`facet_grid()` allows you to specify different types of small multiples

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ class)
```

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_grid(year ~ drv)
```

#### Exercises
One of the great things about facets is that they share the axes between the 
different panels. Sometimes this is undesirable though, and the behavior can
be changed with the `scales` argument. Experiment with the different possible
settings in the plot below:

```{r}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  facet_wrap(~ drv)
```

* * *

Usually the space occupied by each panel is equal. This can create problems when
different scales are used. Can you modify the code below so that the y scale differs 
between the panels in the plot. What happens?

```{r}
ggplot(mpg) + 
  geom_bar(aes(y = manufacturer)) + 
  facet_grid(class ~ .)
```

Use the `space` argument in `facet_grid()` to change the plot above so each bar 
has the same width again.

